Model matters with restricted Boltzmann machines

### 2016-12-16 presentations/amherst/slides.Rmd
edits from dr. nordman

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan, Daniel Nordman, and Stephen Vardeman
shortname: Kaplan, et al.
institute: |
    | Iowa State University
    | ajkaplan@iastate.edu
shortinstitute: ajkaplan@iastate.edu
date: |
  | December 20, 2016
  | 
  |
  | Slides available at <http://bit.ly/kaplan-umass>
shortdate: "December 20, 2016"
output: 
  beamer_presentation:
    template: latex/beamer.tex
    includes:
      in_header: latex/front-matter.tex
theme: CambridgeUS
bibliography: latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

@@

1. Discrete or continuous random variables 
2. two layers - one hidden and one visible
3. conditional independence within a layer [@smolensky1986information]

\begin{figure}
\includegraphics[width=.6\linewidth]{images/rbm.png}
\caption{An example RBM, which consists of two layers. Hidden nodes are indicated by white circles and the visible nodes are indicated by blue circles}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\linewidth]{images/visibles_image.pdf}
\caption{Image classification using a RBM. On the left, each image pixel comprises a node in the visible layer, $\mathcal{V}$. On the right, the output of the RBM is used to create features which are then passed to a supervised learning algorithm.}
\end{figure}

@@

- $\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})$ represents the states of the visible and hidden nodes in a RBM 
- Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$
    - Two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$. 
- A parametric form for probabilities

    \begin{align} 
    \label{eqn:pmf} 
    f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} \end{align}

@@

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}
\begin{itemize}
\item Stacking layers of RBMs in a deep architecture
\item Proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" (Salakhutdinov and Hinton 2009, pp. 450).
\item Treating a hidden layer of one RBM as the visible layer in a second RBM, etc.
\end{itemize}
\end{column}
\hfill
\begin{column}{.48\textwidth}
\begin{figure}
\includegraphics[width=\linewidth]{images/deep_rbm.png}
\caption{Three layer deep Boltzmann machine, with visible-to-hidden and hidden-to-hidden connections but no within-layer connections.}
\end{figure}
\end{column}
\end{columns}

@@

1. The model properties are largely unexplored in the literature and 
2. The commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]

@@

1. Provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and 
2. Explore the possibility of a rigorous fitting methodology

@@

> A model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].

@@

\begin{definition}[Model Degeneracy]
There is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{C}^{\nh + \nv}$, by the model.
\end{definition}

@@

- Concept of model deficiency related to *instability* for a class of exponential families of distributions [@schweinberger2011instability]
- For the RBM, consider how a data model might be expanded to incorporate more visibles
    - Necessary to grow the number of model parameters in a sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ 
    - May also arbitrarily expand the number of hidden variables used.

@@

S-unstable RBM model sequences are undesirable for several reasons - small changes in data can lead to overly-sensitive changes in probability.

@@

\begin{result}
\label{prop:instab}
Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
\end{result}

If the quantity (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in the data configuration.

@@

- All probability will stack up on mode sets or potentially those few outcomes with the highest probability 
- Proofs of results \ref{prop:instab}-\ref{prop:degen} can be found in [@kaplan2016note]

@@

- Measure: the magnitude of the difference between
    1. Model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and 
    2. Expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$
- If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]|$ is large then the RBM with parameter vector $\boldsymbol \theta$ is *uninterpetable*

@@

\begin{figure}
\includegraphics[width=.55\linewidth]{images/toyhull_top.png}
\includegraphics[width=.35\linewidth]{images/toyhull_top_2.png}
\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for the toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
\end{figure}

@@

- To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to *near-degeneracy*, *instability*, and *uninterpretability*, consider models of small size  
- For $\nh, \nv \in \{1, \dots, 4\}$, sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$
    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to
    2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$
    3. Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points
- Calculate metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. 
- In the case of *near-degeneracy*, classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$

@@

1. Computational concerns
    - Fitting a RBM via maximum likelihood (ML) methods infeasible due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$
        - Ad hoc methods are used, aim to avoid this problem by using stochastic ML 
        - employ a small number of MCMC draws to approximate $\gamma(\boldsymbol \theta)$ 
2. With $\nh$ large enough 
    - Potential to re-create any distribution for the data [@le2008representational; @montufar2011refinements; and @montufar2011expressive]
        - The model for the cell probabilities that has the highest likelihood over  *all possible model classes* is the empirical distribution
        - The RBM model ensures that this empirical distribution can be arbitrarily well approximated
    - When empirical distribution contains empty cells, fitting will chase parameters to $\infty$ in order to zero out the corresponding RBM cell probabilities

@@

- Consider what might be done in a principled manner, testing on $\nv = \nh = 4$ 
- To avoid model impropriety, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to *near-degeneracy*, *instability*, and *uninterpretability*.
    - Shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ 
        1. Specify priors that place low probability on large values of $||\boldsymbol \theta||$
        2. Shrink $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$
- Consider a test case with $\nv = \nh = 4$ and parameters given in in appendix 
    - $\boldsymbol \theta$ chosen as a sampled value from a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy (not near the convex hull of the sufficient statistics) 
    - simulate $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies

@@

2. *A truncated Normal prior (BwTNLV)* 
    - Independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$
        - $\sigma_{interaction} < \sigma_{main}$ 
        - *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively
    - Simulation from the posterior using a geometric adaptive MH step [@zhou2014some]
    - $h_j$ are carried along in the MCMC implementation as latent variables 
3. *A truncated Normal prior and marginalized likelihood (BwTNML)* 
    - Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$
    - Use the truncated Normal priors applied to the marginal probabilities for visible variables

@@

- RBMs are thought to be useful for classification, but in the context of generative statistical models, poor fit due to *near-degeneracy*, *S-instability*, and *uninterpretability*
- Rigorous fitting methodology is difficult due to the dimension of the parameter space & size of the latent variable space
- For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive] 
    - The empirical distribution of a training set is the best fitting model for observed cell data 
    - There can be no "smoothed distribution" achieved in a RBM model of sufficient size with a rigorous likelihood-based method

@@

1. Generalization of instability results [ongoing, see @kaplan2016note]
2. Image classification
    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
    - Decision theoretic based approach to approximating the likelihood ratio test for classification
3. Markov chain Monte Carlo methods for data with Markovian dependence
    - Spatial data
    - Network data

@@

---- 
\begin{center}
\Huge
Appendices
\end{center}

>>>>

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan
shortname: Kaplan, et al.
institute: |
    | Iowa State University
    | ajkaplan@iastate.edu
shortinstitute: ajkaplan@iastate.edu
date: |
  | December 20, 2016
  |
  | Slides available at <http://bit.ly/kaplan-umass>
  | 
  | \footnotesize Joint work with D. Nordman and S. Vardeman
shortdate: "December 20, 2016"
output: 
  beamer_presentation:
    template: latex/beamer.tex
    includes:
      in_header: latex/front-matter.tex
theme: CambridgeUS
bibliography: latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

@@

2. two layers of random variables - one hidden and one visible
3. conditional independence within a layer [@smolensky1986information]

\begin{figure}
\includegraphics[width=.6\linewidth]{images/rbm.png}
\caption{Hidden nodes are indicated by white circles and the visible nodes are indicated by blue circles.}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\linewidth]{images/visibles_image.pdf}
\caption{Image classification using a RBM: each image pixel comprises a node in the visible layer, $\mathcal{V}$ and the output of the RBM is used to create features passed to a supervised learning algorithm.}
\end{figure}

@@

- $\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})$ represents visible and hidden nodes in a RBM 
- Each single "binary" random variable, visible $v_i$ or hidden $h_j$, takes values in a common coding set 
    - $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$. 
- A parametric form for probabilities

    \begin{align*} 
    \label{eqn:pmf} 
    f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} \end{align*}

@@

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}
\begin{itemize}
\item Stacking layers of RBMs in a deep architecture
\item Proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" (Salakhutdinov and Hinton 2009, pp. 450).
\end{itemize}
\end{column}
\hfill
\begin{column}{.48\textwidth}
\begin{figure}
\includegraphics[width=\linewidth]{images/deep_rbm.png}
\caption{Three layer deep Boltzmann machine, with visible-to-hidden and hidden-to-hidden connections but no within-layer connections.}
\end{figure}
\end{column}
\end{columns}

@@

1. The model properties are largely unexplored in the literature
2. The commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]

@@

1. Provide steps toward understanding properties of the model class from the perspective of statistical theory
2. Explore the possibility of a rigorous fitting methodology

@@

> A model should "provide an explanation of the mechanism underlying the observed phenomena" [@box1967discrimination].

@@

\begin{definition}[Model Degeneracy]
A disproportionate amount of probability is placed on only a few elements of the sample space, $\mathcal{C}^{\nh + \nv}$, by the model.
\end{definition}

@@

- Concept of model deficiency related to *instability* for a class of exponential families of distributions [@schweinberger2011instability]
- For the RBM, consider how model incorporates more visibles
    - Model parameters in a longer sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ 
    - May also arbitrarily expand the number of hidden variables used.

@@

S-unstable RBM models are undesirable for several reasons - small changes in data outcomes can lead to overly-sensitive changes in probability.

@@

\begin{result}
\label{prop:instab}
Let $c > 0$ and fix an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
\end{result}

If the $\nv^{-1}ELPR(\boldsymbol\theta_{\nv})$ is too large, then a RBM model will exhibit large probability shifts for very small changes in the data configuration.

@@

- All probability will stack up on mode sets or potentially those few outcomes with the highest probability 
- Proofs found in [@kaplan2016note]

@@

- Model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$
- Expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$
- If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]|$ is large then the RBM with parameter vector $\boldsymbol \theta$ is *uninterpetable*

@@

\begin{figure}
\includegraphics[width=.55\linewidth]{images/toyhull_top.png}
\includegraphics[width=.35\linewidth]{images/toyhull_top_2.png}
\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
\end{figure}

@@

- To explore the effects of RBM parameters $\boldsymbol \theta$ on *near-degeneracy*, *instability*, and *uninterpretability*, consider models of small size  
- For $\nh, \nv \in \{1, \dots, 4\}$, sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$
    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$
    2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$
    3. Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points
- Calculate metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. 
- In the case of *near-degeneracy*, classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$

@@

1. Computational concerns: Fitting a RBM via maximum likelihood (ML) methods infeasible due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$
    - Ad hoc methods used to avoid this problem with stochastic ML 
    - Employ a small number of MCMC draws to approximate $\gamma(\boldsymbol \theta)$ 
2. Model parameterization concerns: With enough hiddens, 
    - Potential to re-create any distribution for the data [@le2008representational; @montufar2011refinements; and @montufar2011expressive]
        - The model for the cell probabilities that has the highest likelihood over  *all possible model classes* is the empirical distribution
        - The RBM model ensures that this empirical distribution can be arbitrarily well approximated
    - When empirical distribution contains empty cells, ML will chase parameters to $\infty$ in order to zero out corresponding RBM cell probabilities

@@

- Consider what might be done in a principled manner, small test
- To avoid model impropriety, avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ leading to *near-degeneracy*, *instability*, and *uninterpretability*.
    - Shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ 
        1. Specify priors that place low probability on large values of $||\boldsymbol \theta||$
        2. Shrink $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$
- Consider a test case with $\nv = \nh = 4$ 
    - $\boldsymbol \theta$ chosen as a sampled value from a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy (not near the convex hull of the sufficient statistics) 
    - simulate $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies

@@

2. *A truncated Normal prior (BwTNLV)* 
    - Independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$
        - $\sigma_{interaction} < \sigma_{main}$ 
        - *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively
    - Simulation from the posterior using a geometric adaptive MH step [@zhou2014some]
    - $h_j$ are carried along in the MCMC implementation as latent variables 
3. *A truncated Normal prior and marginalized likelihood (BwTNML)* 
    - Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$
    - Use the truncated Normal priors applied to the marginal probabilities for visible variables (recall visibles are the observed data, hiddens are not)

@@

- RBMs used for classification, but are concerning as statistical models due to *near-degeneracy*, *S-instability*, and *uninterpretability*
- Rigorous fitting methodology is difficult due to the dimension of the parameter space & size of the latent variable space
- Fitting a RBM model is also questionable as any distribution for the visibles can be approximated arbitrarily well
    - The empirical distribution of visibles is the best fitting model for observed cell data 
    - There can be no "smoothed distribution" achieved in a RBM model of sufficient size with a rigorous likelihood-based method

@@

1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
2. Image classification
    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
    - Decision theoretic based approach to approximating the likelihood ratio test for classification
3. Markov chain Monte Carlo methods for data with Markovian dependence
    - Spatial data
    - Network data

@@
